---
title: "Homework 3"
author: "Mingyin Wang"
date: 2024-10-08
output: github_document
---

```{r, echo=FALSE}
library(tidyverse)
library(patchwork)
library(patchwork)
```
Problem 1

```{r}
```

Problem 2
load clean, and tidy data
```{r}
covar_df = read_csv("data/nhanes_covar.csv", skip = 4) |>
  janitor::clean_names() |>
  mutate(
      sex = recode(sex, "1" = "male", "2" = "female"),
      education = recode(education, "1" = "Less than high school", "2" = "High school equivalent", "3" = "More than high school"), 
    sex = factor(sex), 
    education = factor(education)
    )
clean_covar_df = covar_df |>
  filter(age >= 21) |>
  drop_na()

accel_df = read_csv("data/nhanes_accel.csv") |>
  janitor::clean_names() 

```
merge data frame

```{r}
merged_df =  accel_df |>
  inner_join(clean_covar_df, by = "seqn")
```

create a readable table

```{r}
education_sex_table = merged_df |>
  count(education, sex) |>
  spread(key = sex, value = n, fill = 0)
education_sex_table
```

 visualization of the age distributions for men and women in each education category.     
```{r}
ggplot(merged_df, aes(x = education, y = age, fill = sex)) +
  geom_boxplot() + 
  labs(title = "Age Distribution by Education and Sex", x = "Education Level", y = "Age") +
  theme_minimal() 

```

Create a total activity df
```{r}
tot_act = 
  merged_df |>
  mutate(total_activity = rowSums(select(merged_df, starts_with("min"))))

```

Plot total activity vs. age, comparing men and women by education level

```{r}
ggplot(tot_act, aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~education) +
  labs(title = "Total Activity vs. Age by Education and Sex", x = "Age", y = "Total Activity") +
  theme_minimal()
```

pivot longer the df 
```{r}
day_df = tot_act |>
  pivot_longer(cols= starts_with("min"),
               names_prefix = "min",
               names_to = "minute",
               values_to = "mims") |> 
  mutate(minute = as.numeric(minute)) 

```

Make a three-panel plot that shows the 24-hour activity time courses for each education level
```{r}
ggplot(day_df, aes(x = minute, y = mims, color = sex)) +
  geom_smooth(se = FALSE) +
  facet_wrap(~education) +
  labs(title = "24-Hour Activity Time Course by Education and Sex", x = "Minute of the Day", y = "Average MIMS") +
  theme_minimal()
```



