---
title: "Homework 3"
author: "Mingyin Wang"
date: 2024-10-08
output: github_document
---

```{r, echo=FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)

library(p8105.datasets)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Problem 1

```{r}
data("ny_noaa")
```

clean the data, creating separate variables for year, month, and day and converting `tmax` and `tmin` to numeric. 0 is the most commonly observed value for snowfall
```{r}

ny_noaa %>% 
  count(snow)

ny_noaa <- ny_noaa %>%
  mutate(
    year = year(date),
    month = month(date),
    day = day(date),
    tmax = as.numeric(tmax),
    tmin = as.numeric(tmin)
  )
```




```{r}
ny_noaa %>% 
  group_by(id, year, month) %>% 
  filter(month %in% c(1, 7)) %>% 
  summarize(mean_tmax = mean(tmax, na.rm = TRUE, color = id)) %>% 
  ggplot(aes(x = year, y = mean_tmax, group = id)) + geom_point() + geom_path() +
  facet_grid(~month) +
  labs(title = "Mean monthly temperature for each station across years for January and July")
```
The two-panel plot shows the average maximum temperature in January and July for each station across multiple years. The average maximum temperature in January is consistently lower than in July across all stations and years.


a two-panel plot including (i) a hex plot of `tmax` vs `tmin` and ridge plot
```{r}
hex = 
  ny_noaa %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_hex()

ridge = 
  ny_noaa %>% 
  filter(snow < 100, snow > 0) %>%
  ggplot(aes(x = snow, y = as.factor(year))) + 
  geom_density_ridges()

hex + ridge
```
The hex plot shows that most of the data clusters tightly around the center of the distribution, with some variability. In rare cases, tmax values are lower than tmin, suggesting potential data recording or quality issues.

The ridge plot reveals a multimodal distribution of yearly snowfall, with most stations receiving between 0 and 35 mm of snow. Other group of station receive around 45 mm, and another group that sees nearly 80 mm.  This multimodality likely results from the conversion between measurement systems, as noted in the table of common values.













## Problem 2
load clean, and tidy data
```{r}
covar_df = read_csv("data/nhanes_covar.csv", skip = 4) |>
  janitor::clean_names() |>
  mutate(
      sex = recode(sex, "1" = "male", "2" = "female"),
      education = recode(education, "1" = "Less than high school", "2" = "High school equivalent", "3" = "More than high school"), 
    sex = factor(sex), 
    education = factor(education)
    )
clean_covar_df = covar_df |>
  filter(age >= 21) |>
  drop_na()

accel_df = read_csv("data/nhanes_accel.csv") |>
  janitor::clean_names() 

```
merge data frame

```{r}
merged_df =  accel_df |>
  inner_join(clean_covar_df, by = "seqn")
```

create a readable table

```{r}
education_sex_table = merged_df |>
  count(education, sex) |>
  spread(key = sex, value = n, fill = 0)
education_sex_table
```

 visualization of the age distributions for men and women in each education category.         


```{r}
ggplot(merged_df, aes(x = education, y = age, fill = sex)) +
  geom_boxplot() + 
  labs(title = "Age Distribution by Education and Sex", x = "Education Level", y = "Age") +
  theme_minimal()
```

High school equivalent: The age distribution of females tends to be higher than that of males, with a median age of about 60 years for females and around 55 years for males.

Less than high school: The age distribution for males has a broader spread. Females show a median age of about 60, whereas males have a little bit of lower median.

More than high school: Age distribution is similar for both males and females, with medians over 40. But male's median age is higher than female's. 


Create a total activity df
```{r}
tot_act = 
  merged_df |>
  mutate(total_activity = rowSums(select(merged_df, starts_with("min"))))

```

Plot total activity vs. age, comparing men and women by education level

```{r}
ggplot(tot_act, aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~education) +
  labs(title = "Total Activity vs. Age by Education and Sex", x = "Age", y = "Total Activity") +
  theme_minimal()
```
Across all education levels, total activity tends to decrease with age for both males and females, but the rate of decline varies by education level. Individuals with more education tend to maintain relatively higher levels of activity into older age. The "Less than high school" group shows a sharper decline in activity, especially for females. This suggests that education may play a role in promoting consistent physical activity as people age.

pivot longer the df 
```{r}
day_df = tot_act |>
  pivot_longer(cols= starts_with("min"),
               names_prefix = "min",
               names_to = "minute",
               values_to = "mims") |> 
  mutate(minute = as.numeric(minute)) 

```

Make a three-panel plot that shows the 24-hour activity time courses for each education level
```{r}
ggplot(day_df, aes(x = minute, y = mims, color = sex)) +
  geom_smooth(se = FALSE) +
  facet_wrap(~education) +
  labs(title = "24-Hour Activity Time Course by Education and Sex", x = "Minute of the Day", y = "Average MIMS") +
  theme_minimal()
```
Across all education levels, the activity patterns follow a similar daily rhythm with two peaks: one in the morning and one in the afternoon, followed by a decline in the evening.
The highest levels of activity occur during the daytime, with minimal activity during the night.

High school equivalent: Females show slightly higher activity levels than males, especially during the morning peak, though both follow similar trends.

Less than high school: The activity levels for males and females are almost identical, with both showing similar peaks and declines.

More than high school: Females again show higher activity than males throughout most of the day, particularly during the first peak, though the overall patterns are quite similar.

## Question 3

load the datasets
```{r}
jan_2020 = read_csv("data/citibike/Jan 2020 Citi.csv") |>
  janitor::clean_names() |>
  mutate(year = 2020, month = 1)
jan_2024 = read_csv("data/citibike/Jan 2024 Citi.csv") |>
   janitor::clean_names() |>
  mutate(year = 2024, month = 1)
july_2020 = read_csv("data/citibike/July 2020 Citi.csv") |>
   janitor::clean_names() |>
  mutate(year = 2020, month = 7)
july_2024 = read_csv("data/citibike/July 2024 Citi.csv") |>
   janitor::clean_names() |>
 mutate(year = 2024, month = 7)
```

combine all data 
```{r}
combined_df = bind_rows(jan_2020, jan_2024, july_2020, july_2024)
combined_df
```

Produce a reader-friendly table showing the total number of rides in each combination of year and month separating casual riders and Citi Bike members.

```{r}
rides_summary = combined_df |>
  group_by(year, month, member_casual) |>
  summarize(total_rides = n(), .groups = 'drop') |>
  pivot_wider(names_from = member_casual, values_from = total_rides) |>
  mutate(total_rides_month = rowSums(across(c(casual, member), ~replace_na(., 0))))
rides_summary
```
Both casual and member rides increased significantly from 2020 to 2024.The number of rides in July is much higher than in January for both casual riders and members.More people prefer biking during the warmer months. Also, Citi Bike members take more rides than casual riders.


Make a table showing the 5 most popular starting stations for July 2024; include the number of rides originating from these stations.
```{r}
five_popular_stations = combined_df |>
  filter(year == 2024, month == 7) |>
  group_by(start_station_name) |>
  summarize(n_rides = n()) |>
  arrange(desc(n_rides)) |>
  head(5)
five_popular_stations
```

Make a plot to investigate the effects of day of the week, month, and year on median ride duration. 
```{r}
median_ride_duration = 
  combined_df |>
  group_by(year, month, weekdays) |>
  summarize(median_duration = median(duration, na.rm = TRUE), .groups = "drop")


ggplot(median_ride_duration, aes(x = weekdays, y = median_duration, color = factor(year))) +
  geom_line(aes(group = interaction(year, month)), size = 1) +  
  facet_wrap(~ month) +  
  labs(
    title = "Median Ride Duration by Day of the Week, Month, and Year",
    x = "Day of the Week",
    y = "Median Ride Duration (minutes)",
    color = "Year"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
For both 2020 and 2024, median ride durations are noticeably longer in July than in January, indicating that people likely take longer rides during the warmer month of July. 

For July, in 2020, the median ride duration on weekends is significantly higher than on weekdays, with a peak above 15 minutes. For 2024, the weekend rides are also longer but show a slightly less dramatic spike.

For January, in 2020, the median ride duration on weekends is higher than weekdays, with a peak around 10 minutes. The median ride duration in 2024 is fairly consistent throughout the week. 


For data in 2024, make a figure that shows the impact of month, membership status, and bike type on the distribution of ride duration.
```{r}
df_2024 =
  combined_df |> 
  filter(year == 2024) 

ggplot(df_2024, aes(x = duration, fill = rideable_type)) +
  geom_histogram(binwidth = 7, alpha = .4, position = "identity") +   
  facet_wrap(~ month + member_casual) +
  labs(title = "Distribution of Ride Duration by Month, Membership Status, and Bike Type in 2024",
       x = "Ride Duration (minutes)",
       y = "Count",
       fill = "Bike Type") +
  theme_minimal() 
```

Members have more rides with short ride duration time, which means that they might use Citi Bike more for commuting or routine trips. Electric bikes are more frequently used in general.






